---
title: "Cost-based cluster analysis of Cambaceres settlements"
author: "Facundo Muñoz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cambaceres}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message = FALSE, warning = FALSE}
require(tidyr)
require(dplyr)
require(raster)
require(geoRcb)
require(sp)
require(spatstat)
require(ggplot2)
require(viridis)
require(cbK)
theme_set(theme_bw())
```



## Exploration


```{r read-raster, include = FALSE}
## The locations are LazyLoaded with the package
## If you had yourself a shapefile, for example, you 
## would do something like
# basename <- "indigenous_settlements"
# pts <- rgdal::readOGR(dsn = path, layer = basename)

## The cost surface is provided with the package as a
## separate raster file
dn <- system.file("extdata", package="cbK")
cambaceres_cs <- raster(file.path(dn, "cambaceres_cs"))
```


There seem to be some pixels with outlying cost values (an order of magnitude 
higher). These have been filtered in the surface plot.

```{r exploration, fig.width = 5}
boxplot(getValues(cambaceres_cs), horizontal = TRUE)
```


```{r cost-surface-plot, fig.width = 6, fig.height = 4}
data.frame(
  coordinates(cambaceres_cs),
  cost = getValues(cambaceres_cs)
) %>% 
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = cost)) +
  geom_point(
    data = as.data.frame(indigenous_settlements),
    aes(coords.x1, coords.x2)) +
  scale_fill_viridis() +
  coord_fixed()

```


## Cost-based clustering calculations

```{r cbK-results}
res <- cbK(
  loc = indigenous_settlements,
  cs  = cambaceres_cs,
  directions = 16
)
```



## Multi-Dimensional Scaling

In this case, the 2D MDS representation captures the `r round(res$mds$GOF[1]*100, 1)`
 % of the variability in the cost-based distances.
 

## Ripley's $K$ function

```{r ripleys-k, fig.width=5, fig.height=5}
plot(res$K)
plot(res$env_K)
```


Or, its stabilized transformation: the $L$ function

```{r L-function, fig.width=5, fig.height=5}
plot(res$L)
plot(res$env_L)
```



## Comparison with Euclidean distances

The cost-based distances are an order of magnitude higher than the
Euclidean distances.

```{r dist-compare, fig.width=5, fig.height=5}
data.frame(
  eucl = unclass(res$eucl$dm),
  cb = res$dm[lower.tri(res$dm)]
) %>% 
  ggplot(aes(eucl, cb)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, col = 'darkgray')
```


For reference, we have also computed $K$ and $L$ functions with euclidean coordinates.

```{r euclidean-K-L, fig.width=5, fig.height=5}
plot(res$eucl$K)
plot(res$eucl$env_K)

plot(res$eucl$L)
plot(res$eucl$env_L)

```



## Conclusions from Cabaceres case study

- The cost-based approach reveals a strong depart from homogeneity

