---
title: "Cost-based cluster analysis of Tortosa settlements"
author: "Facundo Muñoz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tortosa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message = FALSE, warning = FALSE}
require(tidyr)
require(dplyr)
require(raster)
require(geoRcb)
require(sp)
require(spatstat)
require(ggplot2)
require(viridis)
require(cbK)
theme_set(theme_bw())
```



# Introduction

This vignette presents the results of the analysis for two contrasted historical periods from the same region. In the first case we provide step by step details while in the second we use the function `cbK()` included in the package, who wraps up all the calculations.


## Cost surface

```{r read-raster, include = FALSE}
## The locations are LazyLoaded with the package
## If you had yourself a shapefile, for example, you 
## would do something like
# basename <- "roman_settlements"
# pts <- rgdal::readOGR(dsn = path, layer = basename)

## The cost surface is provided with the package as a
## separate raster file
dn <- system.file("extdata", package="cbK")
tortosa_cs <- raster(file.path(dn, "tortosa_cs"))
```


The cost surface is relatively high-res with `r round(ncell(tortosa_cs)/1e6, 1)` M
pixels, although more than half of them are `NA`s.

There seem to be some pixels with outlying cost values (an order of magnitude 
higher). These have been filtered in the surface plot.

```{r exploration, fig.width = 5}
with(
    list(x = getValues(tortosa_cs)),
  {
    boxplot(x, horizontal = TRUE)
    print(summary(x[x > 200]))
    print(summary(x[x < 1e3]))
  }
)
```


```{r cost-surface-plot, fig.width = 6, fig.height = 4}
data.frame(
  coordinates(tortosa_cs),
  cost = getValues(tortosa_cs)
) %>%                                                             
  filter(cost < 1e3) %>% 
  ggplot(aes(x, y)) +
  geom_raster(aes(fill = cost)) +
  geom_point(
    data = suppressWarnings(
      bind_rows(
        Roman = as.data.frame(roman_settlements),
        Caliphate = as.data.frame(caliphate_settlements),
        .id = "period"
      )
    ),
    aes(coords.x1, coords.x2, col = period)) +
  scale_fill_viridis() +
  coord_fixed()

```


# Roman imperial period


## Computation of the cost-based distance matrix

We prefer to work in terms of conductivity (i.e. inverse-cost).
We write the raster to disk to keep it off-memory.

```{r conductivity-surface}
cond <- writeRaster(1/tortosa_cs, filename=tempfile())
```

This operation is moderately heavy, since the conductivity surface is a raster 
with 14+ M cells. Here we are computing a sparse matrix of size ~ 14 M $\times$ 
14 M. Using `directions = 16` has been problematic for my desktop PC with 16 Gb 
RAM.

```{r compute-distance-matrix}
dm <- distmatGen(roman_settlements, cond, ret = "obs", directions = 8)
```


## Multi-Dimensional Scaling

```{r MDS}
fit <- cmdscale(dm, k =2, eig = TRUE)
```

In this example, the 2D MDS representation captures the `r round(fit$GOF[1]*100, 1)`
 % of the variability in the cost-based distances.
 

## Ripley's $K$ function

First, we need to set up the fitted coordinates as an object of spatial point
pattern (ppp) class.

```{r setup-point-pattern}
cb_pp <- ppp(
  x = fit$points[, 1],
  y = fit$points[, 2],
  xrange = range(fit$points[, 1]),
  yrange = range(fit$points[, 2])
)

```

Then we can compute Ripley's $K$ function, and a Monte Carlo envelope
for significance testing.

```{r ripleys-k, fig.width=5, fig.height=5}
(K <- Kest(cb_pp))
plot(K)
E <- envelope(cb_pp, Kest, verbose = FALSE)
plot(E)
```


Or, its stabilized transformation: the $L$ function

```{r L-function, fig.width=5, fig.height=5}
L <- Lest(cb_pp)
plot(L)
E <- envelope(cb_pp, Lest, verbose = FALSE)
plot(E)
```



## Comparison with Euclidean distances

The cost-based distances are an order of magnitude higher than the
Euclidean distances.

```{r dist-compare, fig.width=5, fig.height=5}
eucl_dm <- dist(coordinates(roman_settlements))
data.frame(
  eucl = unclass(eucl_dm),
  cb = dm[lower.tri(dm)]
) %>% 
  ggplot(aes(eucl, cb)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, col = 'darkgray')
```


We can also compute $K$ and $L$ functions with euclidean coordinates for
reference.

```{r euclidean-K-L, fig.width=5, fig.height=5}
eu_pp <- ppp(
  x = coordinates(roman_settlements)[, 1],
  y = coordinates(roman_settlements)[, 2],
  xrange = range(coordinates(roman_settlements)[, 1]),
  yrange = range(coordinates(roman_settlements)[, 2])
)
K_eu <- Kest(eu_pp)
plot(K_eu)
E_eu <- envelope(eu_pp, Kest, verbose = FALSE)
plot(E_eu)

L_eu <- Lest(eu_pp)
plot(L_eu)
E_eu <- envelope(eu_pp, Lest, verbose = FALSE)
plot(E_eu)

```



## Conclusions from the first example

- The cost-based approach reveals a strong depart from homogeneity



# Caliphate period

The previous computations can be wrapped within one convenience function which
takes a `SpatialPoints` object and a `raster` surface as inputs, and computes all
these results in one step.

This function is provided in this package under the name of `cbK()` (see its 
help page).

```{r caliphate-results}
phase4 <- cbK(
  loc = caliphate_settlements,
  cs  = tortosa_cs,
  directions = 8
)

```



## Multi-Dimensional Scaling

In this case, the 2D MDS representation captures the `r round(phase4$mds$GOF[1]*100, 1)`
 % of the variability in the cost-based distances.
 

## Ripley's $K$ function

```{r phase4-ripleys-k, fig.width=5, fig.height=5}
plot(phase4$K)
plot(phase4$env_K)
```


Or, its stabilized transformation: the $L$ function

```{r phase4-L-function, fig.width=5, fig.height=5}
plot(phase4$L)
plot(phase4$env_L)
```



## Comparison with Euclidean distances

The cost-based distances are an order of magnitude higher than the
Euclidean distances.

```{r phase4-dist-compare, fig.width=5, fig.height=5}
data.frame(
  eucl = unclass(phase4$eucl$dm),
  cb = phase4$dm[lower.tri(phase4$dm)]
) %>% 
  ggplot(aes(eucl, cb)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, col = 'darkgray')
```


For reference, we have also computed $K$ and $L$ functions with euclidean coordinates.

```{r phase4-euclidean-K-L, fig.width=5, fig.height=5}
plot(phase4$eucl$K)
plot(phase4$eucl$env_K)

plot(phase4$eucl$L)
plot(phase4$eucl$env_L)

```


